Information Distribution Depends on Language-Specific Features

### 2019-02-01 cogsci/position-info.tex
through E3

\begin{abstract}
We argue for a reformulation of the Uniform Information Density
hypothesis Levy \& Jaeger (2007) in which the grammatical, phonological
and morphological features of a language determine how information is
distributed throughout utterances. We expand on a method from Yu, Cong,
Liang, \& Liu (2016) to compute entropy based on word position within
utterances as a derivation of information distribution. We analyze adult
and child spoken corpora as well as text corpora drawn from a variety of
languages on Wikipedia. We find that each language has a robust,
characteristic entropy curve across word position, and present evidence
that these entropy curves are in part determined by typological
features. We conclude that speakers may tend towards communicating
information uniformly, but the distribution of information in their
utterances is conditioned on the language they communicate in.

\textbf{Keywords:}
Entropy; information; information theory; communication
\end{abstract}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

The original evidence in Levy \& Jaeger (2007) finds that the insertion
of complementizers in relative clauses in English corresponds to where
neighboring words have high information content. Similarly, Frank \&
Jaeger (2008) argues that contradictions in English such as ``you're''
do not occur when neighboring words are highly informative. The evidence
in favor of UID largely been situation-specific and English-language
driven, while the hypothesis itself has been applied broadly over the
past decade. Applications include determining whether linguistic
alignment takes place (Jaeger \& Snider, 2013), Zipfian word length
distributions (Piantadosi, Tily, \& Gibson, 2011), communication
efficiency (Mahowald, Fedorenko, Piantadosi, \& Gibson, 2013), dialogue
and turn-taking (Xu \& Reitter, 2018) and the significance of ambiguity
in language (Piantadosi, Tily, \& Gibson, 2012), among other research.

However, other recent work has contradicted the UID hypothesis. Similar
to the original Levy \& Jaeger (2007), Zhan \& Levy (2018) focuses on
information distribution at particular points in sentences. Zhan \& Levy
(2018) finds that more information-rich classifiers in Mandarin Chinese
are produced when production of the neighboring noun is difficult, not
when the information content is high. Jain, Singh, Ranjan, Rajkumar, \&
Agarwal (2018) examine word order across spoken sentences in Hindi, a
freer word order language than English, and find that information
density has no significant effect on word order.

Recently, Yu et al. (2016) developed a more direct test of the Uniform
Information Density hypothesis, applying the logic used by Genzel \&
Charniak (2002) to look at the distribution of information \emph{within}
individual sentences. Because people process language
incrementally--using the previous words in a sentence to predict the
words that will come next--the amount of information that a word
contains when seen in isolation should increase over the course of a
sentence. Analyzing a large corpus of written English, they find a
different pattern: Entropy increases over the first few words of an
utterance and then remains constant until the final word where it again
jumps up (see Figure \ref{fig:our_replication}). Yu et al. (2016)
conclude that the Uniform Information Density hypothesis must not hold
for medial words in a sentence.

We extend and generalize Yu et al. (2016) in three ways: we confirm that
this same pattern is found in spoken English--in both adult and child
speakers. We then examine entropy curves cross-linguistically, finding a
diversity of shapes across the world's languages. Finally, we show that
entropy curves are predictable from the structure of individual
languages, namely word order. Taken together, our results suggest a
refinement of the Uniform Information Density hypothesis: speakers may
structure their utterances to optimize information density, but they
must do so under the constraints of the language they use to
communicate.

The UID hypothesis predicts that information transmission rate will tend
towards uniformity in both speech and text. We begin by examining speech
to determine if we obtain the entropy curve shape we expect for spoken
communication. We begin with the same source as Levy \& Jaeger (2007)
and Jaeger (2010), the Switchboard corpus of adult telephone
conversations. We also examine child-adult conversations in the CHILDES
TalkBank (Demuth, Culbertson, \& Alter, 2006; MacWhinney, 2014) corpora
database of spoken adult-child conversations in multiple languages.
Children are not fully developed speakers, so we want to compare the
entropy curve we obtain by computing over the utterances in the CHILDES
corpora to the utterances in the adult-adult Switchboard corpus.
Finally, we extend our cross-linguistic findings on a small number of
languages from the CHILDES experiment to a direct comparison of several
dozen languages in Wikipedia.

\hypertarget{general-methods}{%
\subsection{General methods}\label{general-methods}}

Here we will describe our adaptation of the by-word entropy method in Yu
et al. (2016), which uses the text portion of the British National
Corpus (Clear, 1993). We replicate their results at the end of this
subsection. Given a text or speech corpus divided into individual
utterances, we partition the corpus by utterance length in number of
words. For each word position \(X\) of utterances of length \(k\), we
define \(w\) as a unique word occurring in position \(X\). We further
define \(p(w)\) as the number of times word \(w\) occurs in position
\(X\), divided by the number of total words that occur in position \(X\)
i.e.~the number of sentences of length \(k\). This creates a probability
distribution over the words occurring in position \(X\), and computing
the Shannon entropy \(H(X)\) Shannon (1948) of this probability
distribution gives the positional entropy of position \(X\) in
utterances of length \(k\).

@@

With this measure, we compute the unigram entropy at each position of
sentences of each length within the corpus. The result of this method
can be plotted for each utterance length as an \emph{entropy curve},
which can be visually compared across utterance length to observe the
how the unigram entropy changes across absolute positions in each of the
utterances. Genzel \& Charniak (2002) similarly examine a unigram
entropy measure on sentences, and found that entropy at the sentence
level increases linearly with sentence index within a corpus. UID
applies this uniformity of entropy rate in sentences to all levels of
speech, and so our method obtained from Yu et al. (2016), which examines
text at the word level, should find an affine function at the word
level.

The entropy curves capture individual variation across positions in
utterances of the same length. This allows us to directly observe and
judge the amount of variation in words that appear in an individual
position of a sentence. We can directly compare any two positions within
utterances to determine the amount of uncertainty, and therefore
information, on average contained by words within that position of
utterances. We are applying the same approach as in Genzel \& Charniak
(2002), but within sentences instead of across sentences.

The positional entropy at the beginning of sentences is low, then rises
and plateaus for sentence-medial word positions before dropping slightly
in the second-to-last position and rising again. However, this curve
could simply be a result of the written words in the BNC. We will expand
this analysis and replicate the result with a spoken American English
corpus.

\hypertarget{experiment-1}{%
\section{Experiment 1}\label{experiment-1}}

We replicated the Yu et al. (2016) results and found the same
characteristic three-step distribution as they did for the BNC entropy
curves.

We first used the same source for data as Levy \& Jaeger (2007), the
Switchboard corpus of American telephone conversations Godfrey,
Holliman, \& McDaniel (1992). The Switchboard corpus contains telephone
conversations of American English between strangers, with individual
utterances ranging from one word to several dozen words. We divided the
Switchboard corpus by utterance length and computed the positional
entropy for each position within each sentence length, then plotted the
result as an entropy curve.

@@

Our plots for three representative utterance lengths are below. We found
the same characteristic shape for the American English Switchboard
corpus as in our replication of the Yu et al. (2016) results for the
BNC.

{\centering \includegraphics{figs/read_and_plot_switchboard-1}

\caption[Representative Entropy Curves for the British National Corpus (top) and Switchboard (bottom)]{Representative Entropy Curves for the British National Corpus (top) and Switchboard (bottom). Points average entropies, error bars show 95\% confidence intervals computed by non-parametric bootstrap.}\label{fig:read_and_plot_switchboard}
\end{figure*}
\end{CodeChunk}

The shape we find in these two English data is notable for a number of
reasons, but what we find most interesting is one, the shape of the
distribution does not follow our predictions from UID and two, the
distribution is robust across an unrelated pair of spoken and written
English corpora. This suggests that the entropy curve is characteristic
of the English language at least, and is not a specific effect of
computing positional entropies in the written British English texts in
the BNC. For our next step, we want to expand the positional entropy
analysis to child-adult conversational corpora and languages beyond
English to see if the same three-step distribution appears.

\hypertarget{experiment-2}{%
\section{Experiment 2}\label{experiment-2}}

After replicating the Yu et al. (2016) results using the spoken
conversation corpus from Levy \& Jaeger (2007), we want to determine
whether the entropy curves will remain robust over child-adult
conversational corpora. We also wanted to see whether the curves would
show the same characteristic three-step shape across languages. To
accomplish both of these goals, we used data from the CHILDES Talkbank
system MacWhinney (2014).

The corpora in the CHILDES database consists of short and often
disconnected utterances across hours of recordings, the unigram entropy
measure is unaffected by context or lack thereof in utterances by the
adults and children in the corpora. For our first corpus, we used the
Providence English corpus Demuth et al. (2006). The Providence corpus
recorded interactions between children between 1 and 3 years old and
their parents in the home. We divided the corpus by speaker into child
and non-child categories. We further divided the corpus by utterance
length, so that all sentences of length \(k\) (e.g. \(6\)) were grouped
together. Finally, within each utterance length, we computed the unigram
entropy measure for each position. Computation of positional entropy was
identical to our general method after dividing for utterance length.

We also ran this analysis on Spanish and Mandarin corpora from CHILDES.
Spanish is an Indo-European language like English, possessing similar
grammar, word order and numerous cognate word. Mandarin Chinese is
typologically completed unrelated to English. We used the Shiro corpus
for Spanish Shiro (2000), which contains prompted narratives
individually collected from over a hundred Venezualan schoolchildren,
half from high SES backgrounds and half from low SES backgrounds. We
used the Zhou dinner corpus for Mandarin Chinese Li \& Zhou (2015),
which contains dinner conversations between 5 to 6-year-old children and
their parents collected in Shanghai.

For each corpus, we accessed transcripts of the corpus provided through
the TalkBank system and computed over the Latin alphabet transcriptions
or transliterations of the original transcriptions. For Mandarin, we
used pinyin transliterations of the utterances in the corpus with
demarcated word boundaries. The Chinese characters used for writing
Mandarin do not normally demarcate word boundaries by spacing words
apart, and for normal Chinese writing including spaces between word
boundaries can have a negative effect on reading times Bai, Yan,
Liversedge, Zang, \& Rayner (2008).

We expect that our entropy analysis on the Providence corpus will
produce an identical entropy curve shape to the BNC and Switchboard
curves. We expect similar results to English in the Spanish Shiro
entropy curve, while we expect that our Mandarin corpus will produce a
different entropy curve shape.

@@

The adult and child entropy curves track one another almost identically.
This is surprising because UID would predict that the young children in
the corpora who are not fully developed speakers would have more noise
in their distribution of information rates. This not only indicates a
robustness in the unigram entropy curve across speakers, but also across
ages and addressees. We also observed a robustness across corpora for
the same languages, and a robustness across utterance lengths within the
same corpus's entropy curve. This shows that the concept of an ``entropy
curve'' for a specific language is well-founded when considering speech
data.

We found a distinct three-step distribution for English and Spanish
CHILDES corpora, with a slight dip in the penultimate position of each
sentence. The Mandarin corpus entropy curve, by comparison, has a
noticeably lower positional entropy values in utterance-final positions
than in utterance-penultimate positions. We attribute the penultimate
dip in the English and Spanish entropy curves to the fact that most of
the utterances in the CHILDES English and Spanish corpora we examined
had a determiner such as ``the'' or ``a'' in the second-to-last position
of utterances. The beginnings of utterances in the English and Spanish
CHILDES data were usually pronouns or grammatical subjects, while the
final words were grammatical objects and had a great deal of variation
in the exact word that appeared in the utterance-final position.

These entropy curves are not what would be expected from UID. The
robustness of the three-step distribution for English and its
replication in Spanish do not resemble the affine function we would
expect from UID. The Mandarin entropy curve, which does not at all
resemble either the English/Spanish distribution or the predictions of
UID, suggests that the entropy curve can vary from language to language.
UID predicts that each language should have a similar distribution,
which should approximate an affine function.

We found that all of the English corpora followed the same three-step
entropy curve shape, regardless of whether the corpus contained spoken
or written data. The Spanish corpus we used also followed this same
shape, which we expected. However, Mandarin did not have the same shape,
indicating that the shape of the entropy curves may arise from
cross-linguistic variation. Similar grammatical and other features
between two languages, such as Spanish and English, may produce a
similar entropy curve.

@@

\hypertarget{results-and-analysis-2}{%
\subsection{Results and Analysis}\label{results-and-analysis-2}}

@@

@@

\leavevmode\hypertarget{ref-jaeger2010}{}%
Jaeger, T. F. (2010). Redundancy and reduction: Speakers manage
syntactic information density. \emph{Cognitive Psychology},
\emph{61}(1), 23--62.

@@

>>>>

\begin{abstract}
Although languages vary widely in their structure, all are vehicles for
the transmission of information. Consequently, aspects of speakers' word
choice can be understood through the lens models of optimal
communication. One prediction of these models is that speakers should
keep the density of information constant over the duration of an
utterance (Levy \& Jaeger, 2007). However, different languages have
different coding models that constraint the space of speaker's choices
(e.g.~canonical word order). We build on a mdethod developed by Yu,
Cong, Liang, \& Liu (2016) to analye the entropy curves of natural
language productions across a diverse set of languages and in diverse
written and spoken contexts. We show languages have characteristic
constraints that they impose on speaker's choices that predict
deviations from Uniform Information Density, and that cross-linguistic
variability in these deviations is predictable in part from syntactic
properties of those languages.

\textbf{Keywords:}
Uniform information density; language structure; corpus analysis
\end{abstract}

@@

The original evidence in Levy \& Jaeger (2007) finds that the insertion
of complementizers (e.g. ``that'') in relative clauses in English
corresponds to where neighboring words have high information content.
Similarly, Frank \& Jaeger (2008) argues that contradictions in English
such as ``you're'' do not occur when neighboring words are highly
informative. The evidence in favor of Uniform Information Density
largely been situation-specific and English-language driven, while the
hypothesis itself has been applied broadly over the past decade.
Applications include determining whether linguistic alignment takes
place (Jaeger \& Snider, 2013), Zipfian word length distributions
(Piantadosi, Tily, \& Gibson, 2011), communication efficiency (Mahowald,
Fedorenko, Piantadosi, \& Gibson, 2013), dialogue and turn-taking (Xu \&
Reitter, 2018) and the significance of ambiguity in language
(Piantadosi, Tily, \& Gibson, 2012), among other research.

However, other recent work has contradicted the Uniform Information
Density hypothesis. Similar to the original Levy \& Jaeger (2007), Zhan
\& Levy (2018) focuses on information distribution at particular points
in sentences. Zhan \& Levy (2018) finds that more information-rich
classifiers in Mandarin Chinese are produced when production of the
neighboring noun is difficult, not when the information content is high.
Jain, Singh, Ranjan, Rajkumar, \& Agarwal (2018) examine word order
across spoken sentences in Hindi, a freer word order language than
English, and find that information density has no significant effect on
word order.

Recently, Yu et al. (2016) developed a more direct test of the Uniform
Information Density hypothesis, applying the logic used by Genzel \&
Charniak (2002) to look at the distribution of information \emph{within}
individual sentences. Because people process language
incrementally--using the previous words in a sentence to predict the
words that will come next--the amount of information that a word
contains when seen in isolation should increase over the course of a
sentence (Ferrer-i-Cancho, Debowski, \& Prado Martin, 2013). Analyzing a
large corpus of written English, they find a different pattern: Entropy
increases over the first few words of an utterance and then remains
constant until the final word where it again jumps up (see top of Figure
\ref{fig:read_and_plot_exp1}). Yu et al. (2016) conclude that the
Uniform Information Density hypothesis must not hold for medial words in
a sentence.

We extend and generalize Yu et al. (2016) in three ways: We confirm that
this same pattern is found in spoken English--both in written language,
and in conversational speech between adults and between parents and
their children. Thus, this entropy curve is a robust feature of English
productions. We then examine entropy curves cross-linguistically, and
show that characteristic curves vary across langauge the world's
languages. Finally, we show that this variation is predictable in part
from the structure of individual languages (i.e.~word order). Taken
together, our results suggest a refinement of the Uniform Information
Density hypothesis: speakers may structure their utterances to optimize
information density, but they must do so under the predictable
constraints of their language.

\hypertarget{calculating-entropy-curves}{%
\subsection{Calculating entropy
curves}\label{calculating-entropy-curves}}

In all oll of our studies, we used an adaptation of the by-word entropy
method developed by Yu et al. (2016). Given a text or speech corpus
divided into individual utterances, we partition the corpus by utterance
length in number of words. For each word position \(X\) of utterances of
length \(k\), we define \(w\) as a unique word occurring in position
\(X\). We further define \(p(w)\) as the number of times word \(w\)
occurs in position \(X\), divided by the number of total words that
occur in position \(X\) i.e.~the number of sentences of length \(k\).
This creates a probability distribution over the words occurring in
position \(X\), and computing the Shannon (1948) entropy \(H(X)\) of
this probability distribution gives the positional entropy of position
\(X\) in utterances of length \(k\).

@@

With this measure, we compute the unigram entropy at each position of
sentences of each length within the corpus. The result of this method
can be plotted for each utterance length as an \emph{entropy curve},
which can be visually compared across utterance length to observe the
how the unigram entropy changes across absolute positions in each of the
utterances. Genzel \& Charniak (2002) similarly examine a unigram
entropy measure on sentences, and found that entropy at the sentence
level increases linearly with sentence index within a corpus. Uniform
Information Density applies this uniformity of entropy rate in sentences
to all levels of speech, and so our method obtained from Yu et al.
(2016), which examines text at the word level, should find a
monotonically increasing at the word level.

The entropy curves capture individual variation across positions in
utterances of the same length. This allows us to directly observe and
judge the amount of variation in words that appear in an individual
position of a sentence. We can directly compare any two positions within
utterances to determine the amount of uncertainty, and therefore
information, on average contained by words within that position of
utterances. This method is thus identical in logic to Genzel \& Charniak
(2002), but within sentences instead of across sentences.

\hypertarget{study-1}{%
\section{Study 1}\label{study-1}}

We began by replicating Yu et al.'s (2016) analysis, computing entropy
curves on the British National Corpus--a collection of predominantly
written English (Clear, 1993). We then applied the same method to the
Switchboard corpus--a collection of spoken language (Godfrey, Holliman,
\& McDaniel, 1992). This allows us to ask whether the characteristic
function identified by Yu et al. (2016) is a general feature of English,
or instead a function of written language.

\hypertarget{data-and-analysis}{%
\section{Data and Analysis}\label{data-and-analysis}}

The British National Corpus consists of predominantly (90\%) written
language documents collected in the early 1980s and 1990s across a
variety of genres (scientific articles, newspapers, fiction, etc). It
also contains a small collection of spoken language. All together, it
contains \(\sim\) 100 million tokens. The Switchboard corpus is a
collection of \(\sim\) 2,400 telephone conversations between
unacquainted adults prompted with subjects of conversation. Switchboard
is the corpus used in Levy and Jaeger's (2007) original demonstration of
the Uniform Information Density Hypothesis.

For each corpus, we computed entropy curves using the method described
above for all sentences from length 4 to 30.

\hypertarget{results}{%
\section{Results}\label{results}}

The entropy curves computed in both the British National Corpus and
Switchboard were remarkably consistent both across corpora and across
the range of sentence lenghts we analyzed (Figure
\ref{fig:read_and_plot_exp1}). Confirming Yu et al.'s (2016) findings,
we find that The positional entropy at the beginning of sentences is
low, then rises and plateaus for sentence-medial word positions before
dropping slightly in the second-to-last position and rising again.

{\centering \includegraphics{figs/read_and_plot_exp1-1}

\caption[Representative Entropy Curves for the British National Corpus (top) and Switchboard (bottom)]{Representative Entropy Curves for the British National Corpus (top) and Switchboard (bottom). Points average entropies, error bars show 95\% confidence intervals computed by non-parametric bootstrap.}\label{fig:read_and_plot_exp1}
\end{figure*}
\end{CodeChunk}

The shape of positional entropy curves we find in these two corpora is
notable for two reasons: (1) it does not follow our predictions from
Uniform Information Density, and (2) the distribution is robust across
written and spoken Engilsh. This suggests that the entropy curve is
characteristic of the English language. We next asked whether this shape
is a feature even of children's speech, and also whether it varies
cross-linguistically.

\hypertarget{study-2}{%
\section{Study 2}\label{study-2}}

To understand how robust this entropy curve is, we turned to
conversational speech from parent-child interactions. If we find the
same shape even in children's productions, we have even stronger
evidence that the 3-step curve is a charcateristic feature of English.
We thus turned to the Child Language Data Exchange System (CHILDES), a
collection of transcripts of parent-child interactions (MacWhinney,
2014).

\hypertarget{data-and-analysis-1}{%
\subsection{Data and Analysis}\label{data-and-analysis-1}}

We analyzed three seperate corpora in CHILDES: The Providence Corpus
(Demuth, Culbertson, \& Alter, 2006), The Shiro corpus (Shiro, 2000),
and the Zhou Dinner Corpus (Li \& Zhou, 2015). The Providence corpus
consists of conversations between six 1--3-year-old American English
speaking children and their parents recorded in their homes. Providence
corpus recorded interactions between children between 1 and 3 years old
and their parents in the home. The Shiro Corpus consists of prompted
Spanish-language narratives individually collected from over a hundred
Venezualan schoolchildren, half from high SES backgrounds and half from
low SES backgrounds. The Zhou Dinner Corpus contains dinner
conversations between 5 to 6-year-old Mandarin speaking children and
their parents collected in Shanghai. Spanish is an Indo-European
language like English, possessing similar grammar, word order and
numerous cognate word. In contrast, Mandarin Chinese is typologically
completed unrelated to English.

We accessed the transcripts from each corpus using \texttt{childesr}, an
R-interface to a database formatted version of CHILDES (Sanchez et al.,
in press). We divided all utterances from each transcript into those
produced by target child, and those produced by all other speakers. We
then applied the same entropy curve method described above. For
Mandarin, we used pinyin transliterations of the utterances in the
corpus with demarcated word boundaries. The Chinese characters used for
writing Mandarin do not normally demarcate word boundaries by spacing
words apart, and for normal Chinese writing including spaces between
word boundaries can have a negative effect on reading times (Bai, Yan,
Liversedge, Zang, \& Rayner, 2008).

@@

@@

Across corpora, we found similar entropy curve shapes for adults and
children, but distinct shapes for each language. Figure
\ref{fig:plot_childes} shows representative curves across corpora, but
shapes were robust across the full range utterances we analyzed. We
found a distinct three-step distribution for English and Spanish CHILDES
corpora, with a slight dip in the penultimate position of each sentence.
The Mandarin corpus entropy curve, by comparison, has a noticeably lower
positional entropy values in utterance-final positions than in
utterance-penultimate positions.

These results present two important pieces of information. First, our
analysis of the Providence corpus broadly replicates the shape we found
in both the British National Corpus and in Switchboard. Thus, the
entropy curve of English appears not just in adult-adult conversation,
but even in speech produced by parents to their children, and speech
produced by very young children to their parents. This suggests that it
is a highly robust feature of English langauge, as it structures the
productions of even pre-schooled aged children.

Second, the Entropy curves of English, Spanish, and Mandarin were not
identical. None of these shapes resembled the monotonically increasing
function predicted naively from Uniform Information Density, but also
Mandarin was quite different from both English and Spanish. This
suggests that the entropy curve can vary from language to language. One
possibility is that this variation arises from typological features of
languages, such as characteristic word order. We noticed, for instance,
a relatively high density of determiners in the penultimate position of
English and Spanish utterances, which could account for the penultimate
dip in those languages. In our final analysis, we explored this
possibility directly, analyzing a large set of Wikipedia corpora from
diverse languages, and asking whether variability in their entropy
curves was related to variability in their syntactic structure.

@@

@@

\leavevmode\hypertarget{ref-ferrer-i-cancho2013}{}%
Ferrer-i-Cancho, R., Debowski, L., \& Prado Martin, F. M. del. (2013).
Constant conditional entropy and related hypotheses. \emph{Journal of
Statistical Mechanics: Theory and Experiment}, \emph{2013}(07), L07001.

@@

@@

\leavevmode\hypertarget{ref-sanchez2019}{}%
Sanchez, A., Meylan, S. C., Braginsky, M., MacDonald, K. E., Yurovsky,
D., \& Frank, M. (in press). Childes-db: A flexible and reproducible
interface to the child language data exchange system. \emph{Behavior
Research Methods}.
